```{r setup, include = FALSE}

```

```{r libraries}
require(plyr)
require(ggplot2)
require(knitr)
```

# Tutorial 2

## Results from shape area estimation experiment

The results (from the whole class) of the area estimation experiment from Tutorial 1 are in the following data table and box plot.

```{r data}
x <- read.csv("./tutorial.csv")
kable(x)

kable(ddply(x, .(shape), summarize
      , n = length(shape)
      , mean = mean(estimate)
      , sd = sd(estimate)))
```

```{r analysis1}
ggplot(x, aes(shape, estimate)) + geom_boxplot()

x.lm <- lm(estimate ~ shape, data = x)
plot(x.lm, which = 1:2)

kable(anova(lm(residuals(x.lm)^2 ~ x$shape))) # Levene's test
```

```{r analysis2}
# remove outliers
x <- x[x$estimate < 100, ]
ggplot(x, aes(shape, estimate)) + geom_boxplot()

x.lm <- lm(estimate ~ shape, data = x)
plot(x.lm, which = 1:2)

kable(anova(lm(residuals(x.lm)^2 ~ x$shape))) # Levene's test
```

1. Look at the data table above.  Comment on any unusual features of the data.  Do you think the ANOVA assumptions will be satisfied?

2. Go through the SAS output: interpret and comment.  Include answers to these questions:
   a) There is one analysis on pages 3 to 5 and another on pages 6 to 8.  How do they differ?  What motivated the second analysis?
   b) Where is the median for the triangles in the boxplot on page 6?
   c) Why do the ANOVAs from the first and second analysis give such different results?  Which is preferable?  Are we justified in using the second analysis?
   d) Why do the two Kruskal-Wallis tests show the same result, even though the ANOVA results were so different?

3. How would this be written up as a research report using the Assignment Guidelines from Notices 1?

